---
title: "gapcolor"
author: "Brian S Yandell"
date: "October 7, 2015"
output: html_document
---

This Rmd explores colors with gapminder data. It also explores use of functions.

Attach any libraries.

```{r echo=FALSE}
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})
```

Read gapminder data.

```{r echo=FALSE}
gap <- read.delim("http://tiny.cc/gapminder", sep="\t")
```

Pull out asia. Be sure to drop country levels no longer used.

```{r}
asia <- gap %>% filter(continent == "Asia") %>% 
  select(country, year, lifeExp) %>% droplevels
```

First plot to look.

```{r}
ggplot(asia, aes(x=year, y=lifeExp, color=country)) + geom_line(lwd=1)
```

## Change colors to Jenny Bryan's pallette.

Get Jenny's colors.

```{r}
## http://stat545-ubc.github.io/block019_enforce-color-scheme.html
gdURL <- "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderCountryColors.txt"
countryColors <- read.delim(file = gdURL, as.is = 3) # protect color
jColors <- countryColors$color
names(jColors) <- countryColors$country
```

Now scale color manually.

```{r}
ggplot(asia, aes(x=year, y=lifeExp, color=country)) +
  geom_line(lwd=1) + theme(legend.position = "none") +
  scale_color_manual(values = jColors)
```

## Reorder countries by `lifeExp` in 1952.

```{r}
ord <- asia %>% filter(year == 1952) %>% select(lifeExp) %>% order
asia$country <- ordered(asia$country, 
                        levels = levels(asia$country)[ord])
ggplot(asia, aes(x=year, y=lifeExp, color=country)) + geom_line(lwd=1)
```

Now let's make this a function so we could do on other continents.

```{r}
country.order <- function(gap, cont.name, year.to.order) {
  ## Pull my continent.
  my.cont <- gap %>% filter(continent == cont.name) %>% 
    select(country, year, lifeExp) %>% droplevels
  ## Get numerical order using year.to.order
  ord <- my.cont %>% filter(year == year.to.order) %>% 
    select(lifeExp) %>% order
  ## Note  use of factor inside levels to restrist to countries present.
  my.cont$country <- ordered(my.cont$country, 
                             levels = levels(my.cont$country)[ord])
  ## Return data frame.
  my.cont
}
```

Now try using on Asia with year 1952.

```{r}
asia <- country.order(gap, "Asia", 1952)
ggplot(asia, aes(x=year, y=lifeExp, color=country)) + geom_line(lwd=1)
```

## Another more elegant way using reorder.

Here is a simpler way to do it.

```{r}
Asia <- gap %>% filter(continent == "Asia") %>% 
  select(country, year, lifeExp) %>% droplevels
myfun <- function(x) x[1]
ggplot(Asia, aes(x=year, y=lifeExp, 
                 color=reorder(country, lifeExp, myfun))) +
  geom_line(lwd=1)
```

Here it is all in one line.

```{r}
ggplot(gap %>% filter(continent == "Asia") %>% 
         select(country, year, lifeExp) %>% droplevels, 
       aes(x=year, y=lifeExp,
           color=reorder(country, lifeExp, function(x) x[1]))) +
  geom_line(lwd=1)
```

Now make a function. This one returns the plot

```{r}
country.reorder <- function(gap, cont.name, year.to.order) {
  years <- unique(gap$year)
  year.to.order <- match(year.to.order, years)
  myfun <- function(x) x[year.to.order]
  ggplot(gap %>%                           ## pull out continent.
           filter(continent == cont.name) %>% 
           select(country, year, lifeExp) %>% droplevels, 
         aes(x=year, y=lifeExp,              ## reorder by lifeExp
             color=reorder(country, lifeExp, myfun))
       ) +
  geom_line(lwd=1)
}
```

Now plot as you wish.

```{r}
country.reorder(gap, "Asia", 1952)
```

## Some error checking.

What is missing? What if year.to.date is not in range?

```{r}
country.reorder <- function(gap, cont.name, year.to.order) {
  years <- unique(gap$year)
  year.to.order <- match(year.to.order, years)
  
  ## Put some checks in.
  stopifnot(!is.na(year.to.order))
  stopifnot(cont.name %in% levels(gap$continent))
  
  myfun <- function(x) x[year.to.order]
  ggplot(gap %>%                           ## pull out continent.
           filter(continent == cont.name) %>% 
           select(country, year, lifeExp), 
         aes(x=year, y=lifeExp,              ## reorder by lifeExp
             color=reorder(country, lifeExp, myfun))
       ) +
  geom_line(lwd=1)
}
```

Try it out. The following generate errors.

```
country.reorder(gap, "asia", 1952)
country.reorder(gap, "Asia", 0)
```

## Adding grey scale as an option to function.

This uses a grey scale based on the year.to.order. Here we also allow year.to.order to be a function (such as `median`) rather than a particular year.

```{r}
country.grey <- function(gap, cont.name, year.to.order,
                            color.range = c(0.2,0.8)) {
  
  ## Put some checks in.
  stopifnot(cont.name %in% levels(gap$continent))

  ## Get fancier on order -- let it be a function.
  if(is.function(year.to.order))
    myfun <- year.to.order
  else {
    years <- unique(gap$year)
    year.to.order <- match(year.to.order, years)
    stopifnot(!is.na(year.to.order))
    myfun <- function(x) x[year.to.order]
  }

  ggplot(gap %>%                           ## pull out continent.
           filter(continent == cont.name) %>% 
           select(country, year, lifeExp) %>% droplevels, 
         aes(x=year, y=lifeExp,              ## reorder by lifeExp
             color=reorder(country, lifeExp, myfun))
       ) +
    geom_line(lwd=1) + theme(legend.position = "none") +
    scale_color_grey(start = max(color.range), end = min(color.range))
}
```

Grey scale on Asia using 1952.

```{r}
country.grey(gap, "Asia", 1952)
```

Grey scale on Americas using median.

```{r}
country.grey(gap, "Americas", median)
```

You can play with colors, but there are many challenges with picking color pallettes that work well. Often less is more according to the pros.