---
title: "gapfinal"
author: "Brian S Yandell"
date: "September 17, 2015"
output: html_document
---

Attach any libraries.

```{r echo=FALSE}
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})
```

Read gapminder data.

```{r echo=FALSE}
gap <- read.delim("http://tiny.cc/gapminder", sep="\t")
```

Here is a simple table -- compare `head(gap)` with `tbl_df(gap)`.

``{r}
head(gap)
tbl_df(gap)
```

Now use `dplyr` pure functions `pipe`, `filter`, `select` and `arrange` to focus on variables plotted below for US and China.

```{r}
gap %>%
  filter(country %in% c("China","United States")) %>%
  select(year, country, gdpPercap, lifeExp) %>%
  arrange(year) %>%
  tbl_df
```

Reshape data using `spread` (from `tidyr`) to compare `lifeExp` of two countries. Notice that I `mutate` (from `dplyr`) to change the order of US and China and change the label on `US`.

```{r}
gap %>%
  filter(country %in% c("China","United States")) %>%
  mutate(country = ordered(country, 
    levels=c("United States","China"),
    labels=c("US","China"))) %>%
  select(year, country, lifeExp) %>%
  spread(country, lifeExp)
```

Create `pp` as the basic plot I want in three different ways. Klunky old way (as in gapfinal.Rmd):

```{r}
pp1 <- ggplot(gap[gap$country %in% c("China","United States"),], aes(x=log10(gdpPercap), y=log10(lifeExp), color=country)) + geom_path()
```

More elegant use of `filter` pure function:

```{r}
pp2 <- ggplot(filter(gap, country %in% c("China","United States")),
  aes(x=log10(gdpPercap), 
    y=log10(lifeExp), color=country)) + 
  geom_path()
```

Even more elegant use of pipes:

```{r}
pp3 <- gap %>% 
  filter(country %in% c("China","United States")) %>%
  ggplot(
    aes(x=log10(gdpPercap), 
      y=log10(lifeExp), color=country)) + 
    geom_path()
```

Show these three (code hidden--advanced topic using package `grid`).

```{r, echo=FALSE,fig.height=6.5,fig.width=10}
suppressPackageStartupMessages(library(grid))
pushViewport(viewport(layout = grid.layout(3,1)))
print(pp1, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(pp2, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
print(pp3, vp = viewport(layout.pos.row = 3, layout.pos.col = 1))
```

Now pretty picture with explanatory labels and sized open points.

```{r}
pp3 + geom_point(aes(size=log10(pop)), shape=1) +
  xlab("log10 of GDP Per Capita") + 
  ylab("log10 of Life Expectancy") +
  ggtitle("Gapminder for China and US")
```
